Bee visitation, pollination services, and plant yield in commodity and hybrid seed canola

### 2021-12-01 Manuscript/main.tex
Started working on manuscript

\subsection{Data collection}
From June through August of 2014 and 2015, we surveyed 29 commodity canola fields (14 in 2014, 15 in 2015) near Beaverlodge, Alberta and 31 fields (17 in 2014, 14 in 2015) near Lethbridge, Alberta.
Commodity canola fields were selected based on the proximity of honey bee apiaries and site access. 
28 of the 60 fields were stocked with honey bees (mean: 0.6 hives/hectare, SD: 0.58) while 32 fields were unstocked. %range: 0.03 - 3.1 Apiaries were stationed at the edges of fields, often near to shrubs or tree-lines to act as a wind break.
14 of 31 of the fields near Lethbridge were watered using central-pivot irrigation systems.
% We assumed similar numbers of workers per colony at both localities.
% Growers were asked for canola variety information, but there was not enough replication to test for difference between varieties.
% Fields with central-pivot irrigation were mostly circular, but dryland (non-irrigated) fields were usually square or rectangular.

%Maps + climate graphs needed for publication
During 2015 and 2016, we also surveyed 35 hybrid seed canola fields (15 in 2015, 20 in 2016) near Lethbridge, Alberta, from June through August of each year.
Seed fields were all stocked with honey bee hives at a rate of 3.6 hives/ha, with apiaries stationed in the corners of fields, and were also stocked with leafcutter bees in shelters throughout the field.
Shelters used either ``standard" or ``double"  leafcutter bee stocking ($\approx$50,000 vs. 100,000 cocoons/ha), and shelters were placed at either ``standard" or ``double" densities (1.3 vs. 2.6 shelters/ha) in each field.
Double stocking rates were only used in fields with double shelter density. %, creating three combinations of leafcutter bee stocking: ``standard", ``double shelter", and ``double shelter and double stocking".
All seed fields had central-pivot irrigation, and were mostly circular or semi-circular in shape.
%Woudick 5,6 = 64 shelters (single), Woudick 2 = 113 shelters (double + bees), Woudick 3 = 128 shelters Graham 1 = 135 shelters

@@

In commodity fields, we established plots at 5, 20, 100, and 500m along a transect into the field, starting at the field margin closest to the set of honey bee hives (271 total plots). %used a Garmin\texttrademark eTrex20 GPS, 
In unstocked fields, we began the transect at a field edge close to potential sources of natural pollinating insects, such as forests or grasslands, but observed very few wild pollinators (Table \ref{tab:propVisitors}).
In seed canola fields, bays of ``male" and ``female" plants were planted in 1- and 6-m wide bays, respectively, and leafcutter bee shelters were located in the female bays (Figure \ref{fig:seedfieldPhoto}, \ref{fig:seedfieldPlots}).
We established plots at 5, 20, 100 (250 in 2016), and 400m into the field along a transect from the nearest set of honey bee hives.
We made paired observations in adjacent male and female plots at each distance from the edge of the field (Figure \ref{fig:seedfieldPlots}).
At the edge and centre of each field (5m and 400m), we established another plot in the centre of the female bay to examine the effect of distance away from the male bay.
We also used plot-level leafcutter and honey bee visitation data from Waytes (\citeyear{waytesMsc}, and unpublished; same years and locations) to more accurately estimate the effect of distance from shelter on visitation rates (647 total plots).
We used a Nikon\texttrademark Laser 800S Rangefinder to measure the distance from each plot to the nearest shelter.
All surveys occurred on fair-weather days (median air temperature: 24.5$^\circ$C., range: 17--33) with no rain and as minimal wind as possible (less than 30km/hr).
Mean air temperature was similar between years in Grande Prairie ($\approx$ 22$^\circ$C), but Lethbridge was warmer in 2015 (23 vs. 28$^\circ$C).
Plots were marked with coloured stakes driven into the soil to locate them later in the season. %Full paper should have a table of climate variables. Windspeed needs sorting out (unit problems?)

Within each 1m$^2$ plot, we recorded the number of insect contacting the stigmas and anthers of flowers during 10 minutes (5 min for seed fields during 2015), as well as their taxonomic identity.
We counted visitable flowers in each plot, where ``visitable" was defined as an opened flower whose petals had not yet started to dehisce, and whose style had not elongated more than 3-4 millimeters beyond the anthers.
To assess pollen deposition at the plot level, we collected stigmas from five random open, receptive flowers at each plot, mounted them in fuchsin gel \citep{beattie1971} on depression slides, and counted the pollen on each stigma using a Leica\texttrademark DME 13595 light microscope under 100x magnification (1294 commodity and 1050 seed canola stigmas).
Each field was visited once during the main canola bloom (late June - late July), and again before harvest (mid - late August) to collect plant samples.
We collected three plants and recorded the density of plants per m$^2$ at the same plot where we conducted the visitation observations (789 commodity canola plants, 582 seed canola plants).
After drying the plants at 50$^\circ$C, we weighed the entire plant, counted mature pods, and counted the number of remaining flower pedicels to estimate total flower production.
We randomly sampled five pods from each plant, and counted and weighed the seeds from each pod, (3872 commodity canola pods,  2885 seed canola pods).
Finally, we threshed all the pods for each plant by hand, winnowed them using an air separator, and weighed the total mass of cleaned seeds.

\subsection{Analysis}
To examine how yield is influenced by visitation, we used a piecewise structural equation model framework (pSEM, \citealp{shipley2009}).
Structural equation models are comprised of a set of underlying linear models arranged in a causal network, and provides a framework for testing hypotheses about a complex system \citep{grace2012,lefcheck2015}.
Starting with a simple model (visitation $\rightarrow$ pollen deposition $\rightarrow$ fruit production $\leftarrow$ plant resources), we built more complex models that related measured variables within a causal framework (Figures \ref{fig:commoditySEM} and \ref{fig:seedSEM}).
These models allowed us to infer the relative strength of the main terms (e.g. pollen deposition, plant size) while controlling for other confounding factors (e.g. distance into field).
Fruit set can suppress future vegetative growth and flower production \citep{stephenson1981}, but SEMs do not allow for causal loops \citep{grace2012}, so we included a path from fruit set to flower production.
The direction of the path could plausibly be reversed (whereby large number of flowers reduce the proportion of fruit set), but we considered only the first scenario, as \citet{sabbahi2006} showed that the converse process is more likely (low fruit set causes increased flower production).

The underlying models of each pSEM were specified as generalized linear mixed-effects models (GLMMs, see Equation sets S1 and S2 for model specifications).
We used hierarchical models in order to include processes that occur at different levels; for example, seed size-number trade-offs were modelled at the pod level, whereas pollen deposition was modeled at the flower level.
Normal random effects were used to model variance between fields, within-field (pollen, flower survival, seed count \& size), and within-plot (seed count \& size); if the lowest-level variance term was inestimable, we fit the model without it.
Continuous responses were modeled using log-normal (plant density, plant size), square-root normal (flower density), and exponential-normal distributions (weight per seed), while counts were modeled as a negative-binomial (honey bee visits, pollen grains per stigma, flowers per plant, seeds per pod) or zero-inflated negative-binomial distribution (leafcutter bee visits) with a log-link function.
Visitation models used a $log$(time) term with a slope of 1 (``offset" variable) to account for differences in observation times.
Fruit set (fraction of flowers that matured into pods) was modeled using a beta-binomial distribution with a logit-link function.
Interactions between predictors were evident in honey bee visitation rate (Year:Locality interaction) and plant density (Irrigation:Year) in commodity canola, and for leafcutter bee visitation (Tent Stocking:Distance from edge) in seed canola, and were included in each model (represented as Y-shaped path in Figures \ref{fig:commoditySEM} and \ref{fig:seedSEM}). 
Finally, we created bivariate partial-effect plots of relevant terms within each model to aid in visualising each relationship.

@@

%RC: REMIND READER OF THE EXPECTATIONS, AND WHETHER HYPOTHESES WERE SUPPORTED/REJECTED, RATHER THAN JUST STATING THE RESULTS
%SH: 1) SET UP LOGIC MORE CLEARLY, FOLLOW IT THROUGH IN SAME WAY AS THE REST OF THE PAPER
%2) BE CONSISTENT IN WORDS/PHRASES
%3) BREAK UP PARAGRAPHS & PROVIDE MORE TRANSITION SENTENCES
%4) TALK ABOUT THE REAL WORLD AS MUCH AS POSSIBLE, RATHER THAN VARIABLES

@@

%OVERALL SUMMARY FROM SEMS

>>>>

\subsection{Data collection}
From June through August of 2014 and 2015, we surveyed 29 commodity canola fields (14 in 2014, 15 in 2015) near Beaverlodge, Alberta and 31 fields (17 in 2014, 14 in 2015) near Lethbridge, Alberta.
Commodity canola fields were selected based on the proximity of honey bee apiaries and site access. 
28 of the 60 fields were stocked with honey bees at the corner or side of the field (mean: 0.6 hives/hectare, SD: 0.58) while 32 fields were unstocked. %range: 0.03 - 3.1 Apiaries were stationed at the edges of fields, often near to shrubs or tree-lines to act as a wind break.
14 of 31 of the fields near Lethbridge were watered using central-pivot irrigation systems.
% Growers were asked for canola variety information, but there was not enough replication to test for difference between varieties. 
% Fields with central-pivot irrigation were mostly circular, but dryland (non-irrigated) fields were usually square or rectangular.

%Maps + climate graphs needed for publication
During 2015 and 2016, we also surveyed 35 hybrid seed canola fields (15 in 2015, 20 in 2016) near Lethbridge, Alberta, from June through August of each year.
Seed fields were all stocked with honey bee hives at a rate of 3.6 hives/ha, with apiaries stationed in the corners of fields, and were also stocked with leafcutter bees in shelters throughout the field.
Shelters used either ``standard" or ``double"  leafcutter bee stocking ($\approx$50,000 vs. 100,000 cocoons/ha), and shelters were placed at either ``standard" or ``double" densities (1.3 vs. 2.6 shelters/ha) in each field.
Double stocking rates were only used in fields with double shelter density, creating three combinations of leafcutter bee stocking: ``standard", ``double shelter", and ``double shelter and double stocking".
All seed fields had central-pivot irrigation, and were mostly circular or semi-circular in shape.
%Woudick 5,6 = 64 shelters (single), Woudick 2 = 113 shelters (double + bees), Woudick 3 = 128 shelters Graham 1 = 135 shelters

@@

In commodity fields, we established 1 m$^2$ plots at 5, 20, 100, and 500m along a transect into the field, starting at the field margin closest to the set of honey bee hives (271 total plots). %used a Garmin\texttrademark eTrex20 GPS, 
In unstocked fields, we began the transect at a field edge close to potential sources of natural pollinating insects, such as forests or grasslands, but observed very few wild pollinators (Table \ref{tab:propVisitors}).
In seed canola fields, bays of ``male" and ``female" plants are typically planted in 1- and 6-m wide bays, respectively, and leafcutter bee shelters were located in the female bays (Figure \ref{fig:seedfieldPhoto}, \ref{fig:seedfieldPlots}).
We established plots at 5, 20, 100 (250 in 2016), and 400m into the field along a transect from the nearest set of honey bee hives, using pairs of plots at the edge of adjacent male and female bays at each distance (Figure \ref{fig:seedfieldPlots}).
Bee visitation can vary within female bays (Alfonso Parra, pers. comm.), so at the 5m and 400m plot in each field, we established another plot in the centre of the female bay to examine the effect of distance away from the male bay.
We used a Nikon\texttrademark Laser 800S Rangefinder to measure the distance from each plot to the nearest shelter.
All surveys occurred on fair-weather days (median air temperature: 24.5$^\circ$C., range: 17--33) with no rain and as minimal wind as possible (less than 30km/hr).
Mean air temperature was similar between years in Grande Prairie ($\approx$ 22$^\circ$C), but Lethbridge was warmer in 2015 (23 vs. 28$^\circ$C).
Finally, we incorporated plot-level visitation data from Waytes (\citeyear{waytesMsc}; same years and locations) to more accurately estimate the effect of distance from shelter on visitation rates (647 total plots).
%Plots were marked with coloured stakes driven into the soil to locate them later in the season. %Full paper should have a table of climate variables. Windspeed needs sorting out (unit problems?)

Insect visitation and pollination data was collected once during the main canola bloom at each field (late June - late July), and plants were collected before harvest (mid - late August).
We recorded the number of insect visits that contacted the stigmas and anthers of flowers during 10 minutes of observation (5 min for seed fields during 2015), recorded the taxonomic group of the insect visitors, and counted the number of open, visitable flowers in each plot (petals had not yet started to dehisce, style had not elongated more than 3-4 millimeters beyond the anthers).
To assess pollen deposition, we collected stigmas from five random open flowers at each plot, mounted them in fuchsin gel \citep{beattie1971} on depression slides, and counted the pollen on each stigma using a Leica\texttrademark DME 13595 light microscope under 100x magnification (1294 commodity and 1050 seed canola stigmas).
At the end of the growing season, we collected three plants from the same plot and recorded the density of plants per m$^2$ (789 commodity canola plants, 582 seed canola plants).
After drying the plants, we weighed the entire plant, counted mature pods, and counted the number of flower pedicels to estimate total flower production. % at 50$^\circ$C
We randomly sampled five pods from each plant, and counted and weighed the seeds from each pod (3872 commodity canola pods, 2885 seed canola pods), then we threshed all the pods for each plant by hand, winnowed them using an air separator, and weighed the total mass of cleaned seeds.

\subsection{Analysis}
To examine how yield is influenced by visitation, we used a piecewise structural equation model framework (pSEM, \citealp{shipley2009}).
Structural equation models are comprised of a set of underlying linear models arranged in a causal network, and provides a framework for testing hypotheses about a complex system \citep{grace2012, lefcheck2015}.
Starting with a simple model (visitation $\rightarrow$ pollen deposition $\rightarrow$ fruit production $\leftarrow$ plant resources), we built more complex models that related measured variables within a causal framework (Figures \ref{fig:commoditySEM} and \ref{fig:seedSEM}).
These models allowed us to infer the relative strength of the main terms (e.g. pollen deposition, plant size) while controlling for other confounding factors (e.g. distance into field).
Fruit set can suppress future vegetative growth and flower production \citep{stephenson1981}, but SEMs do not allow for causal loops \citep{grace2012}, so we included a path from fruit set to flower production.
The direction of the path could plausibly be reversed (where large number of flowers reduce the proportion of fruit set), but we considered only the first scenario, as \citet{sabbahi2006} showed that the converse process is more likely (low fruit set causes increased flower production).

The underlying models of each pSEM were specified as generalized linear mixed-effects models (GLMMs, see Equation sets S1 and S2 for model specifications).
We used hierarchical (random effects) models to model processes that occur at different levels; for example, seed size-number trade-offs were modelled at the pod level, whereas pollen deposition was modeled at the flower level.
Random intercepts were used to model variance between fields, within-field (pollen, flower survival, seed count \& size), and within-plot (seed count \& size).
Continuous responses were modeled using log-normal (plant density, plant size), square-root normal (flower density), and exponential-normal distributions (weight per seed), while counts were modeled as a negative-binomial (honey bee visits, pollen grains per stigma, flowers per plant, seeds per pod) or zero-inflated negative-binomial distribution (leafcutter bee visits) with a log-link function.
Visitation models used a $log$(time) term with a slope of 1 (``offset" variable) to account for differences in observation times.
Fruit set (fraction of flowers that matured into pods) was modeled using a beta-binomial distribution with a logit-link function.
Interactions between predictors were evident in honey bee visitation rate (Year:Locality interaction) and plant density (Irrigation:Year) in commodity canola, and for leafcutter bee visitation (Tent Stocking:Distance from edge) in seed canola, and were included in each model (represented as Y-shaped path in Figures \ref{fig:commoditySEM} and \ref{fig:seedSEM}). 
%Finally, we created bivariate partial-effect plots of relevant terms within each model to aid in visualising each relationship.

@@

@@
